---
title: "ISQA_8080 Assignment 1"
author: "Eric Troudt"
date: "September 13, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# __Question 1__ 

```{r Question 1, include=FALSE}

#load libraries
library(caret)
library(ggfortify)
library(ISLR)

# load dataset and set trainControl
carSeat_data <- Carseats
trCtrl <- trainControl(method = 'none')
```


## _Generate multiple regression model for car seat data_
```{r Multple reg model}
lm_carSeat <- train(Sales ~ Price + Urban + US, data = carSeat_data, method = 'lm', trControl = trCtrl)
summary(lm_carSeat)
```

## _How are qualitative predictors encoded by R?_

Urban variable encoding:
```{r Urban contrast}
contrasts(carSeat_data$Urban)
```

US variable encoding: 
```{r US contrast}
contrasts(carSeat_data$US)
```

## _Generate model without Urban variable_
```{r Multple reg model without Urban}
lm_carSeat_2 <- train(Sales ~ Price + US, data = carSeat_data, method = 'lm', trControl = trCtrl)
summary(lm_carSeat_2)
```


## _What is the Root Mean Square Error for each model?_

The RMSE for the 1st regression model (all variables) is: `r RMSE(pred = lm_carSeat$finalModel$fitted.values, carSeat_data[['Sales']])`  

The RMSE for the 2nd regression model (without Urban variable) is: `r RMSE(pred = lm_carSeat_2$finalModel$fitted.values, carSeat_data[['Sales']])`


## _Create diagnostic plots for 2nd regression model_
```{r Diagnostic plots for 2nd reg model}
autoplot(lm_carSeat_2$finalModel)
```

## _Generate 95% confidence interval_
```{r Conf interval}
confint(lm_carSeat_2$finalModel)
```

## _Predict sales for given test data_

The predicted sales when price is 120 will be: `r predict(lm_carSeat_2, newdata = data.frame(Price=120, Urban='Yes', US='Yes'))`


# __Question 2__
## _Create simulated dataset with predictor variables x1 and x2_
```{r Question 2, include=FALSE}

# load addtl libraries
library(ggplot2)

# Create simulated data set
set.seed(1)
x1 <- runif(100)
x2 <- 0.5 * x1 + rnorm(100)/10
y <- 2 + 2 * x1 + 0.3 * x2 + rnorm(100)

# generate dataframe out of each list of values
x_combined <- data.frame(x1, x2, y)
```


## _Determine extent of correlation amongst x1 and x2_

The correlation coefficient for x1 and x2 is: `r cor(x_combined$x1, x_combined$x2, method = 'pearson')`  


```{r Correlation for x1, x2}
ggplot(data = x_combined, aes(x = x1, y = x2)) + geom_point() + xlab("x1") + ylab ("x2") + geom_smooth(method='lm')
```


## _Generate multiple regression model for y_
```{r Multiple regression model for y}
y_lm <- train(y ~ x1 + x2, data = x_combined, method = 'lm', trControl = trCtrl)
summary(y_lm)
```


## _Generate multiple regression model for y using only x1_
```{r Multiple regression model with only x1}
y_lm_1 <- train(y ~ x1, data = x_combined, method = 'lm', trControl = trCtrl)
summary(y_lm_1)
```


## _Generate multiple regression model for y using only x2_
```{r Multiple regression model with only x2}
y_lm_2 <- train(y ~ x2, data = x_combined, method = 'lm', trControl = trCtrl)
summary(y_lm_2)
```


## _Modify simulated dataset with an additional observation_
```{r Modified dataset, include=FALSE}
x_combined[nrow(x_combined) + 1,] <- c(0.1, 0.8, 6) 
```

number of rows: `r nrow(x_combined)`, with new observation: `r x_combined[101,]`

## _For modified dataset, generate multiple regression model for y_
```{r Multiple regression model for y using modified dataset}
y_lm_mod <- train(y ~ x1 + x2, data = x_combined, method = 'lm', trControl = trCtrl)
summary(y_lm_mod)
```

## _For modified dataset, generate multiple regression model for y using only x1_
```{r Multiple regression model with only x1 using modified dataset}
y_lm_mod_1 <- train(y ~ x1, data = x_combined, method = 'lm', trControl = trCtrl)
summary(y_lm_mod_1)
```

## _For modified dataset, generate multiple regression model for y using only x2_
```{r Multiple regression model with only x2 using modified dataset}
y_lm_mod_2 <- train(y ~ x2, data = x_combined, method = 'lm', trControl = trCtrl)
summary(y_lm_mod_2)
```

## _Create diagnostic plots for new models_

__Modified model (using all variables)__   
   
```{r diagnostic plot of mod model}
autoplot(y_lm_mod$finalModel)
```
   
__2nd Modified model (only x1)__   
  
```{r diagnostic plot of mod model with only x1}
autoplot(y_lm_mod_1$finalModel)
```
   
__3rd Modified model (only x2)__  
  
```{r diagnostic plot of mod model with only x2}
autoplot(y_lm_mod_2$finalModel)
```






